\documentclass[nojss,noheadings]{jss}

\usepackage{amsmath,amssymb,amsfonts}
\usepackage{bm}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{enumitem}
\raggedbottom

\newcommand{\R}{\mathbb{R}}
\newcommand{\Rp}{\mathbb{R}_{+}}
\newcommand{\Ind}{\mathbb{I}}
\newcommand{\dd}{\mathrm{d}}
\renewcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\mathbb{V}}
\newcommand{\Cov}{\mathbb{C}\mathrm{ov}}
\newcommand{\tr}{\operatorname{tr}}
\newcommand{\diag}{\operatorname{diag}}
\newcommand{\logit}{\operatorname{logit}}
\newcommand{\expit}{\operatorname{expit}}
\newcommand{\vect}[1]{\bm{#1}}
\newcommand{\mat}[1]{\mathbf{#1}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\TNp}{\mathcal{N}^{+}}
\newcommand{\IG}{\operatorname{IG}}
\newcommand{\GIG}{\operatorname{GIG}}
\newcommand{\Exp}{\operatorname{Exp}}
\newcommand{\AL}{\operatorname{AL}}
\newcommand{\exAL}{\operatorname{exAL}}
\newcommand{\KL}{\operatorname{KL}}
\newcommand{\CRPS}{\operatorname{CRPS}}
\newcommand{\pplc}{\operatorname{pplc}}
\newcommand{\elbo}{\mathcal{L}}
\newcommand{\pigamma}{\pi_\gamma}
\renewcommand{\theequation}{S\arabic{equation}}
\renewcommand{\thetable}{S\arabic{table}}
\renewcommand{\thesection}{S\arabic{section}}
\renewcommand{\thesubsection}{S\arabic{section}.\arabic{subsection}}
\renewcommand{\thesubsubsection}{S\arabic{section}.\arabic{subsection}.\arabic{subsubsection}}

\author{Antonio Aguirre \\ University of California Santa Cruz \And
        Raquel Barata \\ University of California Santa Cruz \AND
        Raquel Prado \\ University of California Santa Cruz \And
        Bruno Sans\'{o} \\ University of California Santa Cruz}
\title{Supplementary Material for ``\pkg{exdqlm}: An \proglang{R} Package for Estimation and Analysis of Flexible Dynamic Quantile Linear Models''}

\Plainauthor{Antonio Aguirre, Raquel Barata, Raquel Prado, Bruno Sanso}
\Plaintitle{Supplementary Material for exdqlm: An R Package for Estimation and Analysis of Flexible Dynamic Quantile Linear Models}
\Shorttitle{Supplementary material for \pkg{exdqlm}}

\Abstract{This supplement derives the posterior targets, full conditional
distributions, variational updates, Laplace-Delta approximations, and numerical
blocks used by the algorithms in \pkg{exdqlm}. It covers dynamic AL/DQLM and
exAL/exDQLM state-space models, transfer-function state augmentation, static
AL/exAL regression, ridge and regularized horseshoe normal-scale (RHS-NS)
shrinkage priors, forecasting, diagnostics, and posterior-predictive synthesis.
The notation follows the main article, and each algorithmic block is linked to
the corresponding exported package function.}

\Keywords{Bayesian quantile regression, dynamic linear models, asymmetric
Laplace likelihood, extended asymmetric Laplace likelihood, MCMC, variational
Bayes}
\Plainkeywords{Bayesian quantile regression, dynamic linear models, asymmetric
Laplace likelihood, extended asymmetric Laplace likelihood, MCMC, variational
Bayes}

\Address{
  Antonio Aguirre and Raquel Barata\\
  Department of Statistics\\
  University of California Santa Cruz\\
  Santa Cruz, CA 95064\\
  E-mail: \email{jaguir26@ucsc.edu}, \email{raquel.a.barata@gmail.com}\\
}

\begin{document}

\section{Scope and relation to the main article}
\label{sec:supp_scope}

The main article describes the user-facing workflow for \pkg{exdqlm}: dynamic
exDQLM and DQLM fitting, transfer-function models, static exAL regression,
forecasting, diagnostics, and posterior-predictive synthesis. This supplement
gives the corresponding posterior targets, full conditional distributions,
variational updates, Laplace-Delta approximations, ELBO components, and
numerical routines used by the package. The derivations build on the AL
mixture representation of \citet{kozumi2011gibbs}, dynamic quantile linear
models \citep{gonccalves2017dynamic}, the exAL/exDQLM formulation
\citep{yan2025family,me}, FFBS for Gaussian state-space models
\citep{carter1994gibbs,fruhwirth1994data}, nonconjugate variational
approximations \citep{wang2013nonconjugatevb}, and the RHS-NS prior
\citep{nishimura2023shrunken}.

Table~\ref{tab:supp_article_crosswalk} gives the main navigation map. The
model sections derive the fitting targets and updates, Section~\ref{sec:numerical_blocks}
collects reusable numerical identities, and Table~\ref{tab:supp_algorithm_map}
links the exported functions to the corresponding algorithmic blocks.

The notation follows the main article. In particular, $\mat{F}_t$ denotes the
observation vector, $\mat{G}_t$ the state evolution matrix, $\mat{W}_t$ the
evolution covariance, $\vect{\theta}_t$ the dynamic state, $p_0$ the target
quantile level, and $\gamma$ the exAL skewness parameter. Dimensions are stated
when each object is introduced.

\begin{table}[!htb]
\centering
\begin{tabularx}{\textwidth}{@{}>{\raggedright\arraybackslash}p{0.25\textwidth}
>{\raggedright\arraybackslash}p{0.20\textwidth}
>{\raggedright\arraybackslash}p{0.16\textwidth}
>{\raggedright\arraybackslash}X@{}}
\toprule
Main article topic & Main article location & Supplement location & Package functions \\
\midrule
exAL mixture and exDQLM hierarchy & Equations (1)--(6) & S2--S4 & \code{exdqlmMCMC()}, \code{exdqlmLDVB()} \\
AL/DQLM special case & Section 2.5 & S3 & \code{dqlm.ind = TRUE}, \code{fix.gamma = TRUE} \\
Discount-factor evolution & Section 2.3 & S3.1, S4.1 & \code{df}, \code{dim.df} \\
Transfer-function exDQLM & Equations (7)--(10) & S5 & \code{exdqlmTransferMCMC()}, \code{exdqlmTransferLDVB()} \\
Static AL/exAL regression & Section 2.5, Example 4 & S6--S7 & \code{exalStaticMCMC()}, \code{exalStaticLDVB()} \\
Forecasting & Equation (15) & S8.1 & \code{exdqlmForecast()} \\
Diagnostics & Equations (11)--(14) & S8.2 & \code{exdqlmDiagnostics()}, \code{exalStaticDiagnostics()} \\
Posterior-predictive synthesis & Example 1 & S8.3 & \code{quantileSynthesis()} \\
Numerical blocks & Section 2.2 & S9 & Slice sampler, Laplace-Delta, ELBO blocks \\
\bottomrule
\end{tabularx}
\caption{Crosswalk between the main article, this supplement, and the
exported \pkg{exdqlm} functions.}
\label{tab:supp_article_crosswalk}
\end{table}

\section{Notation and distributional conventions}

Let $p_0\in(0,1)$ be the target quantile. The response is scalar. In dynamic
models, observations are indexed by $t=1,\ldots,T$; in static regression,
observations are indexed by $i=1,\ldots,n$. For any positive definite matrix
$A$, $|A|$ denotes its determinant.

\subsection{Notation contract}

The symbols $p_\gamma$, $A_\gamma$, $B_\gamma$, and $C_\gamma$ are deterministic
functions of $(p_0,\gamma)$, whereas $\pigamma(\gamma)$ denotes the prior
density for $\gamma$ on $(L,U)$. Generic densities are denoted by $p(\cdot)$
only when no confusion with the exAL quantile map can arise. In variational
sections, expectations without an explicit subscript are taken under the
current mean-field distribution $q$. The symbol $\xi$ is reserved for the
RHS-NS half-Cauchy auxiliary variable; the transformed coordinate for $\gamma$
in Laplace-Delta calculations is denoted by $z_\gamma$.

\begin{table}[!htb]
\centering
\begin{tabularx}{\textwidth}{@{}lX@{}}
\toprule
Symbol & Meaning \\
\midrule
$p_0$ & Target quantile level. \\
$p_\gamma=p(p_0,\gamma)$ & Internal exAL quantile map defined in the main article. \\
$A_{p_0},B_{p_0}$ & AL constants $A(p_0)$ and $B(p_0)$. \\
$A_\gamma,B_\gamma,C_\gamma$ & exAL constants evaluated at $p_\gamma$. \\
$L,U$ & Lower and upper admissible bounds for $\gamma$. \\
$\pigamma(\gamma)$ & Prior density for $\gamma$ on $(L,U)$. \\
$\vect{\theta}_t,\mat{F}_t,\mat{G}_t,\mat{W}_t$ & Dynamic state, observation vector, evolution matrix, and evolution covariance. \\
$v_t,s_t$ & exAL latent variables in the dynamic model. \\
$\vect{\beta},\mat{X}$ & Static regression coefficients and design matrix. \\
$v_i,s_i$ & exAL latent variables in the static model. \\
$\mathcal{A}$ & Active coefficient set receiving RHS-NS shrinkage. \\
$\elbo(q)$ & Evidence lower bound. \\
\bottomrule
\end{tabularx}
\caption{Core notation used throughout the supplement. The shorthand
$p_\gamma$, $A_\gamma$, $B_\gamma$, and $C_\gamma$ refers to functions of
$(p_0,\gamma)$ defined in the main article.}
\label{tab:supp_notation}
\end{table}

The distributional parameterizations are also fixed throughout:
$\N(m,V)$ uses variance $V$;
\begin{equation}
x\sim\IG(a,b)
\quad\Longleftrightarrow\quad
p(x)=\frac{b^a}{\Gamma(a)}x^{-a-1}\exp(-b/x),\qquad x>0,
\end{equation}
equivalently $x^{-1}\sim\operatorname{Gamma}(a,b)$ under a shape-rate gamma
parameterization. The exponential distribution $\Exp(\text{rate}=1/\sigma)$
has density $\sigma^{-1}\exp(-x/\sigma)$, matching the main article's
mean-$\sigma$ convention for the AL and exAL mixture variables. Finally,
$\TNp(m,V)$ denotes $\N(m,V)$ truncated to $(0,\infty)$. The generalized
inverse Gaussian parameterization is given in \eqref{eq:gig_parameterization}.

The asymmetric Laplace (AL) mixture representation used by the package is
\begin{align}
v &\mid \sigma \sim \Exp(\text{rate}=1/\sigma), \qquad v>0, \\
y \mid \eta,\sigma,v
&\sim \N\{ \eta + A(p_0)v,\ \sigma B(p_0)v \},
\end{align}
where
\begin{equation}
A(p)=\frac{1-2p}{p(1-p)}, \qquad B(p)=\frac{2}{p(1-p)}.
\label{eq:al_AB}
\end{equation}
The AL model is the $\gamma=0$ special case of the exAL formulation below.

The exAL distribution is represented through an additional latent variable
$s\sim\TNp(0,1)$, a standard normal truncated to $(0,\infty)$. For
$\gamma\in(L,U)$, define
\begin{align}
g(\gamma) &= 2\Phi(-|\gamma|)\exp(\gamma^2/2), \\
p_\gamma &= \Ind(\gamma<0) + \frac{p_0-\Ind(\gamma<0)}{g(\gamma)}, \\
A_\gamma &= A(p_\gamma), \qquad
B_\gamma = B(p_\gamma), \qquad
C_\gamma = \{\Ind(\gamma>0)-p_\gamma\}^{-1}.
\label{eq:exal_ABC}
\end{align}
The endpoints $L<0<U$ are the roots $g(L)=1-p_0$ and $g(U)=p_0$. Conditional on
$(v,s,\sigma,\gamma)$,
\begin{equation}
y\mid \eta,\sigma,\gamma,v,s
\sim
\N\left(\eta + C_\gamma\sigma|\gamma|s + A_\gamma v,\ 
\sigma B_\gamma v \right).
\label{eq:exal_aug}
\end{equation}
The package uses an inverse-gamma prior $\sigma\sim\IG(a_\sigma,b_\sigma)$
and a proper truncated Student-$t$ prior $\pigamma(\gamma)$ for $\gamma$ on
$(L,U)$. If $m_\gamma$, $s_\gamma$, and $\nu_\gamma$ denote its location,
scale, and degrees of freedom, then
\begin{equation}
\pigamma(\gamma)=
\frac{
s_\gamma^{-1}t_{\nu_\gamma}\{(\gamma-m_\gamma)/s_\gamma\}
}{
T_{\nu_\gamma}\{(U-m_\gamma)/s_\gamma\}
-T_{\nu_\gamma}\{(L-m_\gamma)/s_\gamma\}
}\Ind(L<\gamma<U),
\label{eq:gamma_prior}
\end{equation}
where $t_\nu(\cdot)$ and $T_\nu(\cdot)$ are the standard Student-$t$ density
and distribution function.

The generalized inverse Gaussian distribution is parameterized as
\begin{equation}
x\sim\GIG(\lambda,\chi,\psi_{\mathrm{GIG}})
\quad\Longleftrightarrow\quad
p(x)\propto x^{\lambda-1}
\exp\left\{-\frac{1}{2}\left(\frac{\chi}{x}+\psi_{\mathrm{GIG}} x\right)\right\},
\qquad x>0.
\label{eq:gig_parameterization}
\end{equation}
For $r\in\R$,
\begin{equation}
\E(x^r)=
\left(\frac{\chi}{\psi_{\mathrm{GIG}}}\right)^{r/2}
\frac{K_{\lambda+r}(\sqrt{\chi\psi_{\mathrm{GIG}}})}
{K_{\lambda}(\sqrt{\chi\psi_{\mathrm{GIG}}})},
\label{eq:gig_moments}
\end{equation}
where $K_\nu(\cdot)$ is the modified Bessel function of the second kind.

\section{Dynamic DQLM under the AL likelihood}
\label{sec:dyn_dqlm}

\subsection{Model hierarchy and full joint}

Let $\vect{\theta}_t\in\R^q$ be the latent state, $\mat{F}_t\in\R^q$ the
observation vector, $\mat{G}_t\in\R^{q\times q}$ the evolution matrix, and
$\mat{W}_t\in\R^{q\times q}$ a positive semidefinite evolution covariance. The
dynamic DQLM is
\begin{align}
\vect{\theta}_0 &\sim \N(\vect{m}_0,\mat{C}_0), \qquad
  \vect{m}_0\in\R^q,\ \mat{C}_0\in\R^{q\times q}, \\
\vect{\theta}_t\mid\vect{\theta}_{t-1}
&\sim \N(\mat{G}_t\vect{\theta}_{t-1},\mat{W}_t), \\
v_t\mid\sigma &\sim \Exp(\text{rate}=1/\sigma), \\
y_t\mid\vect{\theta}_t,\sigma,v_t
&\sim \N(\mat{F}_t^\top\vect{\theta}_t + A_{p_0} v_t,\ \sigma B_{p_0} v_t),
\label{eq:dyn_dqlm_hierarchy}
\end{align}
where $A_{p_0}=A(p_0)$ and $B_{p_0}=B(p_0)$. With
$\vect{v}=(v_1,\ldots,v_T)^\top$ and
$\vect{\theta}_{0:T}=(\vect{\theta}_0,\ldots,\vect{\theta}_T)$, the posterior
target is
\begin{align}
p(\vect{\theta}_{0:T},\sigma,\vect{v}\mid\vect{y})
&\propto
p(\sigma)\,p(\vect{\theta}_0)
\prod_{t=1}^T p(\vect{\theta}_t\mid\vect{\theta}_{t-1})
\prod_{t=1}^T p(v_t\mid\sigma)
\prod_{t=1}^T p(y_t\mid\vect{\theta}_t,\sigma,v_t).
\label{eq:dyn_dqlm_joint}
\end{align}
Equivalently, conditional on $(\sigma,\vect{v})$, the pseudo-observation
$\tilde y_t=y_t-A_{p_0}v_t$ satisfies
\begin{equation}
\tilde y_t=\mat{F}_t^\top\vect{\theta}_t+\epsilon_t,\qquad
\epsilon_t\sim\N(0,\sigma B_{p_0}v_t).
\label{eq:dyn_dqlm_pseudo}
\end{equation}
Thus the conditional state posterior is a Gaussian state-space posterior.
The package implements the evolution variance through the discount-factor
state-space construction used by \code{exdqlmMCMC()} and \code{exdqlmLDVB()};
the notation above writes the induced covariance generically as $\mat{W}_t$.
The derivations condition on the sequence $\mat{W}_{1:T}$ induced by the
discount-factor construction in the main article. Componentwise discounting
changes the block structure of $\mat{W}_t$; once this sequence is fixed within
a Gaussian pseudo-state update, the FFBS and LDVB state updates have the forms
below.

\subsection{MCMC updates}

One Gibbs sweep for the dynamic DQLM uses the following full conditionals.

\paragraph{State path.}
Given $(\sigma,\vect{v})$, sample the state path
$\vect{\theta}_{0:T}$ exactly by applying forward-filtering
backward-sampling (FFBS) to \eqref{eq:dyn_dqlm_pseudo}.

\paragraph{Scale.}
Let $r_t=y_t-\mat{F}_t^\top\vect{\theta}_t$. Then
\begin{equation}
\sigma\mid \vect{\theta}_{0:T},\vect{v},\vect{y}
\sim \IG\left(
a_\sigma+\frac{3T}{2},\
b_\sigma+\sum_{t=1}^T v_t+
\sum_{t=1}^T\frac{(r_t-A_{p_0}v_t)^2}{2B_{p_0}v_t}
\right).
\label{eq:dyn_dqlm_sigma_cond}
\end{equation}

\paragraph{Latent AL variables.}
For $t=1,\ldots,T$,
\begin{equation}
v_t\mid\vect{\theta}_t,\sigma,y_t
\sim
\GIG\left(
\frac{1}{2},\
\frac{r_t^2}{\sigma B_{p_0}},\
\frac{A_{p_0}^2}{\sigma B_{p_0}}+\frac{2}{\sigma}
\right).
\label{eq:dyn_dqlm_v_cond}
\end{equation}

\paragraph{Algorithm S1 (dynamic DQLM MCMC).}
The MCMC kernel leaves the augmented posterior in
\eqref{eq:dyn_dqlm_joint} invariant; posterior summaries are Monte Carlo
estimates from the retained finite chain.
\begin{enumerate}[label=(\arabic*)]
\item Initialize $\vect{\theta}_{0:T}$, $\vect{v}$, and $\sigma$.
\item For $m=1,\ldots,N_{\mathrm{burn}}+N_{\mathrm{mcmc}}$:
  \begin{enumerate}[label=(\alph*)]
  \item sample $\vect{\theta}_{0:T}$ by FFBS using
  \eqref{eq:dyn_dqlm_pseudo};
  \item sample $\sigma$ from \eqref{eq:dyn_dqlm_sigma_cond};
  \item sample each $v_t$ from \eqref{eq:dyn_dqlm_v_cond};
  \item retain draws after burn-in.
  \end{enumerate}
\item Return posterior samples and filtered/smoothed state summaries.
\end{enumerate}

\subsection{Mean-field VB and ELBO}

The dynamic DQLM VB approximation uses
\begin{equation}
q(\vect{\theta}_{0:T},\sigma,\vect{v})
=q(\vect{\theta}_{0:T})q(\sigma)\prod_{t=1}^Tq(v_t),
\label{eq:dyn_dqlm_mf}
\end{equation}
with $q(\sigma)=\IG(a_q,b_q)$ and
$q(v_t)=\GIG(1/2,\chi_t,\psi_t)$. Define
\begin{equation}
\kappa=\E_q(\sigma^{-1})=\frac{a_q}{b_q},\qquad
\ell_t=\E_q(v_t^{-1}),\qquad
\bar v_t=\E_q(v_t).
\end{equation}
The state factor $q(\vect{\theta}_{0:T})$ is a joint Gaussian smoothing
distribution, not a product over time. The latent-variable updates require
only its marginal means and variances.
The state factor is the Gaussian posterior of the auxiliary model
\begin{equation}
y_t^\star=\mat{F}_t^\top\vect{\theta}_t+\epsilon_t^\star,\qquad
\epsilon_t^\star\sim\N(0,R_t^\star),
\qquad
y_t^\star=y_t-\frac{A_{p_0}}{\ell_t},\quad
R_t^\star=\frac{B_{p_0}}{\kappa\ell_t}.
\label{eq:dyn_dqlm_vb_pseudo}
\end{equation}
After Kalman smoothing, write
$q(\vect{\theta}_t)=\N(\vect{m}_{t|T},\mat{C}_{t|T})$ and define
\begin{align}
\bar r_t &= y_t-\mat{F}_t^\top\vect{m}_{t|T},\\
\overline{r_t^2}
&= \bar r_t^2+\mat{F}_t^\top\mat{C}_{t|T}\mat{F}_t.
\end{align}
The remaining coordinate-ascent variational inference (CAVI) updates are
\begin{align}
q(v_t)&=\GIG\left(
\frac{1}{2},\
\frac{\kappa}{B_{p_0}}\overline{r_t^2},\
\kappa\left(2+\frac{A_{p_0}^2}{B_{p_0}}\right)\right), \\
q(\sigma)&=\IG(a_q,b_q), \qquad
a_q=a_\sigma+\frac{3T}{2},\\
b_q&=b_\sigma+\sum_{t=1}^T\bar v_t+
\frac{1}{2B_{p_0}}\sum_{t=1}^T
\left\{\ell_t\overline{r_t^2}-2A_{p_0}\bar r_t+A_{p_0}^2\bar v_t\right\}.
\label{eq:dyn_dqlm_vb_sigma}
\end{align}

\paragraph{Algorithm S2 (dynamic DQLM LDVB).}
\begin{enumerate}[label=(\arabic*)]
\item Initialize the variational parameters for $q(\vect{\theta}_{0:T})$,
$q(\sigma)$, and $q(v_t)$.
\item Repeat until convergence:
  \begin{enumerate}[label=(\alph*)]
  \item update $q(\vect{\theta}_{0:T})$ by Kalman smoothing under
  \eqref{eq:dyn_dqlm_vb_pseudo};
  \item update $q(v_t)$ for $t=1,\ldots,T$;
  \item update $q(\sigma)$ using \eqref{eq:dyn_dqlm_vb_sigma};
  \item monitor moment changes and, when enabled, $\elbo(q)$.
  \end{enumerate}
\item Draw approximate posterior samples from the fitted variational factors
when requested by \code{n.samp}.
\end{enumerate}

The ELBO is
\begin{align}
\elbo(q)
&=
\E_q\{\log p(\vect{y},\vect{\theta}_{0:T},\sigma,\vect{v})\}
-\E_q\{\log q(\vect{\theta}_{0:T},\sigma,\vect{v})\} \\
&=\elbo_\theta+\E_q\{\log p(\sigma)\}
+\sum_{t=1}^T\E_q\{\log p(v_t\mid\sigma)\}
+\sum_{t=1}^T\E_q\{\log p(y_t\mid\vect{\theta}_t,\sigma,v_t)\}\\
&\quad
+H\{q(\sigma)\}+\sum_{t=1}^TH\{q(v_t)\}.
\label{eq:dyn_dqlm_elbo}
\end{align}
The state-path term $\elbo_\theta$ is evaluated from the auxiliary Gaussian
state-space model in \eqref{eq:dyn_dqlm_vb_pseudo} through the Kalman marginal
likelihood identity
\begin{equation}
\elbo_\theta
=\log p(\vect{y}^{\star})
-\E_q\{\log p(\vect{y}^{\star}\mid\vect{\theta}_{0:T})\},
\label{eq:dyn_dqlm_elbo_theta}
\end{equation}
where $p(\vect{y}^{\star})$ is the marginal likelihood of the auxiliary
Gaussian state-space model. This identity is equivalent to
$\E_q\{\log p(\vect{\theta}_{0:T})\}+H\{q(\vect{\theta}_{0:T})\}$ for the
current state update.

The remaining terms in \eqref{eq:dyn_dqlm_elbo} are computable from the
current variational moments. Let
$\bar\ell_\sigma=\E(\log\sigma)$ and
$\bar\ell_{v,t}=\E(\log v_t)$. Then
\begin{align}
\E_q\{\log p(\sigma)\}
&=
a_\sigma\log b_\sigma-\log\Gamma(a_\sigma)
-(a_\sigma+1)\bar\ell_\sigma-b_\sigma\kappa,\\
\sum_{t=1}^T\E_q\{\log p(v_t\mid\sigma)\}
&=
-T\bar\ell_\sigma-\kappa\sum_{t=1}^T\nu_t,\\
\sum_{t=1}^T\E_q\{\log p(y_t\mid\vect{\theta}_t,\sigma,v_t)\}
&=
-\frac{T}{2}\log(2\pi)-\frac{T}{2}\log B_{p_0}
-\frac{T}{2}\bar\ell_\sigma
-\frac{1}{2}\sum_{t=1}^T\bar\ell_{v,t}\nonumber\\
&\quad
-\frac{\kappa}{2B_{p_0}}\sum_{t=1}^T
\left\{\ell_t\overline{r_t^2}-2A_{p_0}\bar r_t+A_{p_0}^2\bar v_t\right\}.
\label{eq:dyn_dqlm_elbo_terms}
\end{align}
The entropy terms are given by \eqref{eq:entropy_ig} for $q(\sigma)$ and by
\eqref{eq:entropy_gig} for the $q(v_t)$ factors.

\section{Dynamic exDQLM under the exAL likelihood}
\label{sec:dyn_exdqlm}

\subsection{Model hierarchy and full joint}

The dynamic exDQLM replaces the AL observation in
\eqref{eq:dyn_dqlm_hierarchy} with the exAL augmentation
\begin{align}
v_t\mid\sigma &\sim \Exp(\text{rate}=1/\sigma), \qquad
s_t\sim\TNp(0,1), \\
y_t\mid\vect{\theta}_t,\sigma,\gamma,v_t,s_t
&\sim
\N\left(
\mat{F}_t^\top\vect{\theta}_t+C_\gamma\sigma|\gamma|s_t+A_\gamma v_t,\
\sigma B_\gamma v_t
\right),
\label{eq:dyn_exdqlm_obs}
\end{align}
with the same state evolution as in Section~\ref{sec:dyn_dqlm}. The complete
posterior target is
\begin{align}
p(\vect{\theta}_{0:T},\vect{v},\vect{s},\sigma,\gamma\mid\vect{y})
&\propto
p(\vect{\theta}_0)\prod_{t=1}^Tp(\vect{\theta}_t\mid\vect{\theta}_{t-1})
\prod_{t=1}^Tp(y_t\mid\vect{\theta}_t,v_t,s_t,\sigma,\gamma) \nonumber\\
&\quad\times
\prod_{t=1}^Tp(v_t\mid\sigma)\prod_{t=1}^Tp(s_t)\,p(\sigma)\pigamma(\gamma).
\label{eq:dyn_exdqlm_joint}
\end{align}
Conditional on $(\vect{v},\vect{s},\sigma,\gamma)$,
\begin{equation}
\tilde y_t=y_t-A_\gamma v_t-C_\gamma\sigma|\gamma|s_t,\qquad
R^y_t=\sigma B_\gamma v_t,
\label{eq:dyn_exdqlm_pseudo}
\end{equation}
which again gives a linear Gaussian state-space model for
$\vect{\theta}_{0:T}$.

\subsection{MCMC updates}

The default package path samples $\sigma$ from its exact conditional
given $\gamma$ and then updates $\gamma$ by bounded univariate slice sampling.
Joint random-walk alternatives are also implemented, but the following target
is the default exact-kernel path.

\paragraph{Latent $v_t$.}
Let
$r_t=y_t-\mat{F}_t^\top\vect{\theta}_t-C_\gamma\sigma|\gamma|s_t$. Then
\begin{equation}
v_t\mid\cdot
\sim
\GIG\left(
\frac{1}{2},\
\frac{r_t^2}{\sigma B_\gamma},\
\frac{A_\gamma^2}{\sigma B_\gamma}+\frac{2}{\sigma}
\right).
\label{eq:dyn_exdqlm_v_cond}
\end{equation}

\paragraph{Latent $s_t$.}
Let $y_t^\circ=y_t-\mat{F}_t^\top\vect{\theta}_t-A_\gamma v_t$ and
$D_\gamma=C_\gamma\sigma|\gamma|$. Then
\begin{equation}
s_t\mid\cdot\sim\TNp(m_{s,t},V_{s,t}),
\qquad
V_{s,t}=\left(1+\frac{D_\gamma^2}{\sigma B_\gamma v_t}\right)^{-1},
\qquad
m_{s,t}=V_{s,t}\frac{D_\gamma y_t^\circ}{\sigma B_\gamma v_t}.
\label{eq:dyn_exdqlm_s_cond}
\end{equation}

\paragraph{State path.}
Given $(\vect{v},\vect{s},\sigma,\gamma)$, sample
$\vect{\theta}_{0:T}$ by FFBS using \eqref{eq:dyn_exdqlm_pseudo}.

\paragraph{Scale.}
For fixed $\gamma$, let $y_t^\circ=y_t-\mat{F}_t^\top\vect{\theta}_t-A_\gamma v_t$
and $c_{\gamma,t}=C_\gamma|\gamma|s_t$. Then
\begin{equation}
\sigma\mid\cdot
\sim
\GIG(\lambda_\sigma,\chi_\sigma,\psi_\sigma),
\label{eq:dyn_exdqlm_sigma_cond}
\end{equation}
where
\begin{align}
\lambda_\sigma &= -\left(a_\sigma+\frac{3T}{2}\right), \\
\chi_\sigma &= 2b_\sigma+2\sum_{t=1}^T v_t+
\sum_{t=1}^T\frac{(y_t^\circ)^2}{B_\gamma v_t},\\
\psi_\sigma &= \sum_{t=1}^T\frac{c_{\gamma,t}^2}{B_\gamma v_t}.
\end{align}

\paragraph{Skewness.}
The conditional kernel for $\gamma$ is
\begin{align}
\pi(\gamma\mid\cdot)
&\propto \pigamma(\gamma)\,
\prod_{t=1}^T(\sigma B_\gamma v_t)^{-1/2}
\exp\left[
-\sum_{t=1}^T
\frac{
\{y_t-\mat{F}_t^\top\vect{\theta}_t-A_\gamma v_t-C_\gamma\sigma|\gamma|s_t\}^2
}{2\sigma B_\gamma v_t}
\right],
\label{eq:dyn_exdqlm_gamma_kernel}
\end{align}
for $\gamma\in(L,U)$. The default sampler applies the bounded slice update in
Section~\ref{sec:slice} directly to this kernel.

\paragraph{Algorithm S3 (dynamic exDQLM MCMC).}
The default $(\sigma,\gamma)$ update consists of an exact GIG update for
$\sigma$ conditional on $\gamma$, followed by a bounded univariate slice update
for $\gamma$. Joint random-walk alternatives are implemented, but they are not
the default kernel summarized here.
\begin{enumerate}[label=(\arabic*)]
\item Initialize $\vect{\theta}_{0:T}$, $\vect{v}$, $\vect{s}$, $\sigma$, and
$\gamma$.
\item For each MCMC iteration:
  \begin{enumerate}[label=(\alph*)]
  \item sample $v_t$ from \eqref{eq:dyn_exdqlm_v_cond}, $t=1,\ldots,T$;
  \item sample $s_t$ from \eqref{eq:dyn_exdqlm_s_cond}, $t=1,\ldots,T$;
  \item sample $\vect{\theta}_{0:T}$ by FFBS using
  \eqref{eq:dyn_exdqlm_pseudo};
  \item sample $\sigma$ from \eqref{eq:dyn_exdqlm_sigma_cond};
  \item sample $\gamma$ on $(L,U)$ using \eqref{eq:dyn_exdqlm_gamma_kernel};
  \item retain draws after burn-in.
  \end{enumerate}
\item Return posterior samples and state summaries.
\end{enumerate}
Setting $\gamma=0$ drops the $s_t$ block and reduces the sampler to the
AL/DQLM kernel in Algorithm S1.

\subsection{Laplace-Delta VB approximation}

The mean-field family is
\begin{equation}
q(\vect{\theta}_{0:T},\vect{v},\vect{s},\sigma,\gamma)
=q(\vect{\theta}_{0:T})\prod_{t=1}^Tq(v_t)\prod_{t=1}^Tq(s_t)\,q(\sigma,\gamma).
\label{eq:dyn_exdqlm_mf}
\end{equation}
The conjugate factors are updated conditionally on moments from
$q(\sigma,\gamma)$. Define
\begin{align}
\kappa_1 &= \E\{1/(\sigma B_\gamma)\},&
\kappa_2 &= \E\{A_\gamma/(\sigma B_\gamma)\},&
\kappa_3 &= \E\{A_\gamma^2/(\sigma B_\gamma)\},\\
\kappa_4 &= \E\{C_\gamma|\gamma|/B_\gamma\},&
\kappa_5 &= \E\{\sigma C_\gamma^2\gamma^2/B_\gamma\},&
\kappa_6 &= \E\{A_\gamma C_\gamma|\gamma|/B_\gamma\}.
\label{eq:dyn_exdqlm_kappa}
\end{align}
Let $m_{1/v,t}=\E(v_t^{-1})$, $m_{s,t}=\E(s_t)$, and
$m_{s^2,t}=\E(s_t^2)$. The state factor is Gaussian and is computed by Kalman
smoothing with observation precision
\begin{equation}
\phi_t=\kappa_1m_{1/v,t}
\end{equation}
and pseudo-response
\begin{equation}
y_t^{\mathrm{VB}}
=
\frac{
y_t\phi_t-\kappa_2-\kappa_4m_{s,t}m_{1/v,t}
}{\phi_t}.
\label{eq:dyn_exdqlm_vb_pseudo}
\end{equation}
If $\eta_t=\mat{F}_t^\top\vect{\theta}_t$, write
$m_{\eta,t}=\E(\eta_t)$, $S_{\eta,t}=\Var(\eta_t)$,
$m_{\Delta,t}=y_t-m_{\eta,t}$, and
$S_{\Delta,t}=m_{\Delta,t}^2+S_{\eta,t}$. Then
\begin{align}
q(v_t)&=\GIG\left(
\frac{1}{2},\
\kappa_1S_{\Delta,t}-2\kappa_4m_{\Delta,t}m_{s,t}+\kappa_5m_{s^2,t},\
\kappa_3+2\E(\sigma^{-1})
\right),\\
q(s_t)&=\TNp(m_{s,t}^{\mathrm{VB}},V_{s,t}^{\mathrm{VB}}),\\
V_{s,t}^{\mathrm{VB}}
&=\left(1+\kappa_5m_{1/v,t}\right)^{-1},\\
m_{s,t}^{\mathrm{VB}}
&=V_{s,t}^{\mathrm{VB}}
\left(\kappa_4m_{\Delta,t}m_{1/v,t}-\kappa_6\right).
\label{eq:dyn_exdqlm_vb_vs}
\end{align}
The nonconjugate factor $q(\sigma,\gamma)$ is approximated by a
Laplace-Delta step on transformed coordinates
\begin{equation}
\ell_\sigma=\log\sigma,\qquad
z_\gamma=\logit\left(\frac{\gamma-L}{U-L}\right).
\label{eq:dyn_exdqlm_ld_transform}
\end{equation}
The transformed objective is the expected complete log-joint plus the
Jacobian,
\begin{align}
h_z(\ell_\sigma,z_\gamma)
&=
\E_{-\sigma,\gamma}\{\log p(\vect{y},\vect{\theta}_{0:T},
\vect{v},\vect{s},\sigma,\gamma)\} \nonumber\\
&\quad
+\ell_\sigma+\log(U-L)+\log r+\log(1-r),
\qquad r=\expit(z_\gamma).
\label{eq:dyn_exdqlm_ld_objective}
\end{align}
The package stores the transformed mode, the approximate covariance matrix,
Delta-method moments, and ELBO components.

\paragraph{Algorithm S4 (dynamic exDQLM LDVB).}
\begin{enumerate}[label=(\arabic*)]
\item Initialize variational parameters for $q(\vect{\theta}_{0:T})$,
$q(v_t)$, $q(s_t)$, and $q(\sigma,\gamma)$.
\item Repeat until convergence:
  \begin{enumerate}[label=(\alph*)]
  \item update $q(\vect{\theta}_{0:T})$ by Kalman smoothing under
  \eqref{eq:dyn_exdqlm_vb_pseudo};
  \item update $q(v_t)$ and $q(s_t)$ using \eqref{eq:dyn_exdqlm_vb_vs};
  \item update $q(\sigma,\gamma)$ by the Laplace-Delta approximation to
  \eqref{eq:dyn_exdqlm_ld_objective};
  \item monitor moment changes and, when enabled, $\elbo(q)$.
  \end{enumerate}
\item Draw approximate posterior samples from the fitted variational factors
when requested by \code{n.samp}.
\end{enumerate}
In the AL special case, the nonconjugate $(\sigma,\gamma)$ block is replaced
by the inverse-gamma $q(\sigma)$ update in Algorithm S2.

\paragraph{ELBO.}
The LDVB ELBO has the block decomposition
\begin{equation}
\elbo(q)=
\elbo_{\mathrm{like}}+\elbo_\theta+\elbo_v+\elbo_s+\elbo_{\sigma\gamma}
+\sum_{t=1}^TH\{q(v_t)\}
+\sum_{t=1}^TH\{q(s_t)\}
+H\{q(\sigma,\gamma)\}.
\label{eq:dyn_exdqlm_elbo}
\end{equation}
The state block is
\begin{equation}
\elbo_\theta
=\E_q\left\{\log p(\vect{\theta}_0)
+\sum_{t=1}^T\log p(\vect{\theta}_t\mid\vect{\theta}_{t-1})\right\}
+H\{q(\vect{\theta}_{0:T})\}.
\label{eq:dyn_exdqlm_elbo_theta}
\end{equation}
As in the AL case, this Gaussian state-space contribution can be evaluated by
the auxiliary Kalman normalizing-constant identity for the current
pseudo-response and variance.

Let
$\bar\ell_\sigma=\E(\log\sigma)$,
$\bar\ell_B=\E(\log B_\gamma)$,
$\bar\ell_{\pi}=\E\{\log\pigamma(\gamma)\}$,
$m_{v,t}=\E(v_t)$, $m_{1/v,t}=\E(v_t^{-1})$,
$m_{\log v,t}=\E(\log v_t)$, $m_{s,t}=\E(s_t)$, and
$m_{s^2,t}=\E(s_t^2)$. Define
\begin{align}
Q_t
&=
\kappa_1S_{\Delta,t}m_{1/v,t}
-2\kappa_2m_{\Delta,t}
+\kappa_3m_{v,t}
-2\kappa_4m_{\Delta,t}m_{s,t}m_{1/v,t} \nonumber\\
&\quad
+\kappa_5m_{s^2,t}m_{1/v,t}
+2\kappa_6m_{s,t}.
\label{eq:dyn_exdqlm_elbo_Q}
\end{align}
Then the likelihood, latent-prior, and nonconjugate-prior blocks are
\begin{align}
\elbo_{\mathrm{like}}
&=
-\frac{T}{2}\log(2\pi)
-\frac{T}{2}(\bar\ell_\sigma+\bar\ell_B)
-\frac{1}{2}\sum_{t=1}^Tm_{\log v,t}
-\frac{1}{2}\sum_{t=1}^TQ_t,\\
\elbo_v
&=
-T\bar\ell_\sigma-\E(\sigma^{-1})\sum_{t=1}^Tm_{v,t},\\
\elbo_s
&=
T\log 2-\frac{T}{2}\log(2\pi)
-\frac{1}{2}\sum_{t=1}^Tm_{s^2,t},\\
\elbo_{\sigma\gamma}
&=
a_\sigma\log b_\sigma-\log\Gamma(a_\sigma)
-(a_\sigma+1)\bar\ell_\sigma
-b_\sigma\E(\sigma^{-1})
+\bar\ell_{\pi}.
\label{eq:dyn_exdqlm_elbo_terms}
\end{align}
The GIG and truncated-normal entropy terms are given by
\eqref{eq:entropy_gig} and \eqref{eq:entropy_tnp}. The
$(\sigma,\gamma)$ entropy and the smooth expectations in the $\kappa_j$,
$\bar\ell_B$, $\bar\ell_\pi$, and $\E(\sigma^{-1})$ terms are evaluated under
the Laplace-Delta approximation in \eqref{eq:entropy_ld} and
\eqref{eq:delta_generic}.

\section{Transfer-function exDQLM by state augmentation}
\label{sec:transfer_function}

Transfer-function inference uses the same AL/exAL likelihood and latent-variable
updates as the dynamic models above. The only change is the state-space
construction. Let $x_t\in\R^k$ be the input vector, $\zeta_t$ the transfer
state, and $\psi_t\in\R^k$ the input-effect state. For fixed transfer rate
$\lambda$, the augmented model uses
\begin{align}
\widetilde{\mat{F}}_t^\top
&=(\mat{F}_t^\top,\ 1,\ 0_k^\top),&
\widetilde{\vect{\theta}}_t^\top
&=(\vect{\theta}_t^\top,\ \zeta_t,\ \psi_t^\top),
\label{eq:tf_aug_state}\\
\widetilde{\mat{G}}_t
&=
\operatorname{blockdiag}\left\{
\mat{G}_t,\,
\begin{pmatrix}
\lambda & x_t^\top\\
0_k & I_k
\end{pmatrix}
\right\},&
\widetilde{\mat{W}}_t
&=\operatorname{blockdiag}\{\mat{W}_t^\theta,\omega_t,W_t^\psi\}.
\label{eq:tf_aug_mats}
\end{align}
The prior for $(\zeta_0,\psi_0^\top)^\top$ is
$\N(m_0^{\mathrm{tf}},C_0^{\mathrm{tf}})$ and is appended to the base
state prior. Conditional on $\lambda$, the transfer-function model is an
ordinary exDQLM on the augmented state $\widetilde{\vect{\theta}}_t$.
No separate transfer-function posterior derivation is needed. Algorithms S3 and
S4 apply after replacing
$(\mat{F}_t,\mat{G}_t,\mat{W}_t,\vect{\theta}_t)$ by
$(\widetilde{\mat{F}}_t,\widetilde{\mat{G}}_t,\widetilde{\mat{W}}_t,
\widetilde{\vect{\theta}}_t)$.

\begin{table}[!htb]
\centering
\begin{tabularx}{0.82\textwidth}{@{}lX@{}}
\toprule
Mathematical object & Package interface \\
\midrule
$x_t$ & \code{X} \\
$\lambda$ & \code{lam} \\
$(\delta_\zeta,\delta_\psi)$ & \code{tf.df} \\
$m_0^{\mathrm{tf}}$ & \code{tf.m0} \\
$C_0^{\mathrm{tf}}$ & \code{tf.C0} \\
median decay time & returned field \code{median.kt} \\
\bottomrule
\end{tabularx}
\caption{Transfer-function quantities and their corresponding package
interface fields.}
\label{tab:supp_tf_inputs}
\end{table}

\paragraph{Algorithm S5 (transfer-function exDQLM).}
\begin{enumerate}[label=(\arabic*)]
\item Construct $\widetilde{\mat{F}}_t$, $\widetilde{\mat{G}}_t$,
$\widetilde{\mat{W}}_t$, and $\widetilde{\vect{\theta}}_t$ from
\eqref{eq:tf_aug_state}--\eqref{eq:tf_aug_mats}.
\item Set the augmented initial prior.
\item Run the selected fitting routine: MCMC uses Algorithm S3 and LDVB uses
Algorithm S4.
\item Return summaries for both the base state and the transfer states.
\end{enumerate}
The transfer-function extension changes only the state-space matrices; the
observation likelihood and latent-variable updates are unchanged.

\section{Static AL and exAL regression with Gaussian priors}
\label{sec:static_ridge}

\subsection{Model hierarchy and full joint}

Let $\mat{X}\in\R^{n\times p}$ have row vectors $\vect{x}_i^\top$, and let
$\vect{\beta}\in\R^p$. The ridge/Gaussian coefficient prior is
\begin{equation}
\vect{\beta}\sim\N(\vect{b}_{\beta,0},\mat{V}_{\beta,0}),\qquad
\vect{b}_{\beta,0}\in\R^p,\quad \mat{V}_{\beta,0}\in\R^{p\times p}.
\label{eq:static_beta_ridge}
\end{equation}
In the package interface, the same Gaussian update is used for
\code{beta_prior = "ridge"}. When \code{beta_prior = "rhs_ns"}, the diagonal
prior precision in this update is replaced by the RHS-NS expected precision
described in Section~\ref{sec:static_rhs}.
\begin{table}[!htb]
\centering
\begin{tabularx}{0.88\textwidth}{@{}lXcc@{}}
\toprule
\texttt{beta\_prior} & Coefficient prior & MCMC & LDVB \\
\midrule
\texttt{ridge} & Gaussian/ridge prior in \eqref{eq:static_beta_ridge}. & yes & yes \\
\texttt{rhs\_ns} & Conjugate RHS-NS prior in Section~\ref{sec:static_rhs}. & yes & yes \\
\bottomrule
\end{tabularx}
\caption{Static coefficient-prior options covered by the supplement.}
\label{tab:supp_static_prior_menu}
\end{table}
Under the exAL likelihood,
\begin{align}
v_i\mid\sigma &\sim \Exp(\text{rate}=1/\sigma),\qquad
s_i\sim\TNp(0,1),\\
y_i\mid\vect{\beta},\sigma,\gamma,v_i,s_i
&\sim
\N\left(
\vect{x}_i^\top\vect{\beta}+C_\gamma\sigma|\gamma|s_i+A_\gamma v_i,\
\sigma B_\gamma v_i
\right).
\label{eq:static_exal_hierarchy}
\end{align}
The static AL model is obtained by setting $\gamma=0$, removing $s_i$, and
using $A_{p_0},B_{p_0}$. The exAL full joint is
\begin{align}
p(\vect{\beta},\sigma,\gamma,\vect{v},\vect{s},\vect{y})
&=
p(\vect{\beta})p(\sigma)\pigamma(\gamma)
\prod_{i=1}^n
p(y_i\mid\vect{\beta},\sigma,\gamma,v_i,s_i)
p(v_i\mid\sigma)p(s_i).
\label{eq:static_exal_joint}
\end{align}

\subsection{MCMC updates}

For fixed $(\sigma,\gamma,\vect{v},\vect{s})$, define
\begin{equation}
y_i^\star=y_i-C_\gamma\sigma|\gamma|s_i-A_\gamma v_i,\qquad
w_i=(\sigma B_\gamma v_i)^{-1}.
\end{equation}
Let $\mat{W}=\diag(w_1,\ldots,w_n)$ and
$\vect{y}^\star=(y_1^\star,\ldots,y_n^\star)^\top$. Then
\begin{align}
\mat{\Sigma}_{\beta\mid\cdot}
&=\left(\mat{V}_{\beta,0}^{-1}+\mat{X}^\top\mat{W}\mat{X}\right)^{-1},\\
\vect{m}_{\beta\mid\cdot}
&=\mat{\Sigma}_{\beta\mid\cdot}
\left(\mat{V}_{\beta,0}^{-1}\vect{b}_{\beta,0}+\mat{X}^\top\mat{W}\vect{y}^\star\right),\\
\vect{\beta}\mid\cdot
&\sim\N(\vect{m}_{\beta\mid\cdot},\mat{\Sigma}_{\beta\mid\cdot}).
\label{eq:static_beta_cond}
\end{align}
Let $D_\gamma=C_\gamma|\gamma|$. The remaining exAL full conditionals are
\begin{align}
s_i\mid\cdot
&\sim \TNp(m_{s,i},V_{s,i}),\\
V_{s,i}
&=\left(1+\frac{\sigma D_\gamma^2}{B_\gamma v_i}\right)^{-1},\\
m_{s,i}
&=V_{s,i}\frac{D_\gamma(y_i-\vect{x}_i^\top\vect{\beta}-A_\gamma v_i)}
{B_\gamma v_i},
\label{eq:static_exal_s_cond}
\end{align}
and, with $r_i=y_i-\vect{x}_i^\top\vect{\beta}-\sigma D_\gamma s_i$,
\begin{equation}
v_i\mid\cdot
\sim
\GIG\left(
\frac{1}{2},\
\frac{r_i^2}{\sigma B_\gamma},\
\frac{A_\gamma^2}{\sigma B_\gamma}+\frac{2}{\sigma}
\right).
\label{eq:static_exal_v_cond}
\end{equation}
For fixed $\gamma$, define $t_i=y_i-\vect{x}_i^\top\vect{\beta}-A_\gamma v_i$.
Then
\begin{equation}
\sigma\mid\cdot
\sim
\GIG\left(
-a_\sigma-\frac{3n}{2},\
2b_\sigma+2\sum_{i=1}^n v_i+\sum_{i=1}^n\frac{t_i^2}{B_\gamma v_i},\
\sum_{i=1}^n\frac{D_\gamma^2s_i^2}{B_\gamma v_i}
\right).
\label{eq:static_exal_sigma_cond}
\end{equation}
The nonstandard $\gamma$ kernel is
\begin{align}
\pi(\gamma\mid\cdot)
&\propto
\pigamma(\gamma)\,[B_\gamma]^{-n/2}
\exp\left\{
\sum_{i=1}^n\frac{D_\gamma s_it_i}{B_\gamma v_i}
-\frac{1}{2\sigma}\sum_{i=1}^n\frac{t_i^2}{B_\gamma v_i}
-\frac{\sigma}{2}\sum_{i=1}^n\frac{D_\gamma^2s_i^2}{B_\gamma v_i}
\right\},
\label{eq:static_exal_gamma_cond}
\end{align}
on $(L,U)$. The package default updates this one-dimensional target by the
bounded slice sampler in Section~\ref{sec:slice}.

For the AL special case, the Gibbs updates reduce to
\begin{align}
\vect{\beta}\mid\cdot
&\sim\N(\vect{m}_{\beta\mid\cdot},\mat{\Sigma}_{\beta\mid\cdot}),\\
\sigma\mid\cdot
&\sim
\IG\left(
a_\sigma+\frac{3n}{2},\
b_\sigma+\sum_{i=1}^n v_i+
\sum_{i=1}^n\frac{(y_i-\vect{x}_i^\top\vect{\beta}-A_{p_0}v_i)^2}{2B_{p_0}v_i}
\right),\\
v_i\mid\cdot
&\sim
\GIG\left(
\frac{1}{2},\
\frac{(y_i-\vect{x}_i^\top\vect{\beta})^2}{\sigma B_{p_0}},\
\frac{A_{p_0}^2}{\sigma B_{p_0}}+\frac{2}{\sigma}
\right).
\label{eq:static_al_mcmc}
\end{align}

\paragraph{Algorithm S6 (static AL/exAL MCMC).}
\begin{enumerate}[label=(\arabic*)]
\item Initialize $\vect{\beta}$, $\vect{v}$, $\vect{s}$, $\sigma$, and
$\gamma$; omit $\vect{s}$ and fix $\gamma=0$ for the AL path.
\item For each MCMC iteration:
  \begin{enumerate}[label=(\alph*)]
  \item sample $\vect{\beta}$ from \eqref{eq:static_beta_cond};
  \item sample $v_i$ for $i=1,\ldots,n$;
  \item for exAL, sample $s_i$ for $i=1,\ldots,n$;
  \item sample $\sigma$;
  \item for exAL, sample $\gamma$ by bounded slice sampling;
  \item if \code{beta_prior = "rhs_ns"}, update the RHS-NS scale block in
  Section~\ref{sec:static_rhs};
  \item retain draws after burn-in.
  \end{enumerate}
\item Return coefficient, latent-variable, and scale posterior samples.
\end{enumerate}

\subsection{Laplace-Delta VB updates and ELBO}

The static exAL variational family is
\begin{equation}
q(\vect{\beta},\vect{v},\vect{s},\sigma,\gamma)
=q(\vect{\beta})\prod_{i=1}^nq(v_i)\prod_{i=1}^nq(s_i)q(\sigma,\gamma).
\label{eq:static_exal_mf}
\end{equation}
The coefficient factor is Gaussian. With the same moment definitions as in
\eqref{eq:dyn_exdqlm_kappa},
\begin{align}
\omega_i&=\kappa_1\E(v_i^{-1}),\\
\eta_i&=
y_i\kappa_1\E(v_i^{-1})
-\kappa_4\E(v_i^{-1})\E(s_i)
-\kappa_2,\\
\mat{\Sigma}_\beta
&=\left(\mat{V}_{\beta,0}^{-1}+\mat{X}^\top\diag(\omega_1,\ldots,\omega_n)\mat{X}\right)^{-1},\\
\vect{m}_\beta
&=\mat{\Sigma}_\beta
\left(\mat{V}_{\beta,0}^{-1}\vect{b}_{\beta,0}+\mat{X}^\top\vect{\eta}\right),
\label{eq:static_exal_vb_beta}
\end{align}
where $\vect{\eta}=(\eta_1,\ldots,\eta_n)^\top$. The $q(v_i)$ and $q(s_i)$
updates are also closed form conditional on moments of $q(\sigma,\gamma)$.
Let
\begin{equation}
\bar r_i=y_i-\vect{x}_i^\top\vect{m}_\beta,\qquad
\overline{r_i^2}=\bar r_i^2+\vect{x}_i^\top\mat{\Sigma}_\beta\vect{x}_i.
\end{equation}
Then
\begin{align}
q(v_i)
&=
\GIG\left(
\frac{1}{2},\
\kappa_1\overline{r_i^2}
-2\kappa_4\bar r_i\E(s_i)
+\kappa_5\E(s_i^2),\
\kappa_3+2\E(\sigma^{-1})
\right),\\
q(s_i)&=\TNp(m_{s,i}^{\mathrm{VB}},V_{s,i}^{\mathrm{VB}}),\\
V_{s,i}^{\mathrm{VB}}
&=\left(1+\kappa_5\E(v_i^{-1})\right)^{-1},\\
m_{s,i}^{\mathrm{VB}}
&=V_{s,i}^{\mathrm{VB}}
\left\{\kappa_4\bar r_i\E(v_i^{-1})-\kappa_6\right\}.
\label{eq:static_exal_vb_vs}
\end{align}
The nonconjugate factor $q(\sigma,\gamma)$ is proportional to
\begin{align}
q(\sigma,\gamma)
&\propto
\pigamma(\gamma)[B_\gamma]^{-n/2}
\sigma^{-(a_\sigma+1+3n/2)}
\exp\left\{
-\frac{1}{2}\left(\frac{\bar c(\gamma)}{\sigma}
+\bar d(\gamma)\sigma\right)+\bar e(\gamma)
\right\},
\label{eq:static_exal_q_sigmagam}
\end{align}
where
\begin{align}
\bar c(\gamma)
&=2b_\sigma+2\sum_{i=1}^n\E(v_i)
+\sum_{i=1}^n
\frac{\E\{(y_i-\vect{x}_i^\top\vect{\beta}-A_\gamma v_i)^2/v_i\}}
{B_\gamma},\\
\bar d(\gamma)
&=\sum_{i=1}^n\frac{C_\gamma^2\gamma^2}{B_\gamma}\E\{s_i^2/v_i\},\\
\bar e(\gamma)
&=\sum_{i=1}^n
\frac{C_\gamma|\gamma|}{B_\gamma}
\E\{s_i(y_i-\vect{x}_i^\top\vect{\beta}-A_\gamma v_i)/v_i\}.
\end{align}
This factor is approximated by the same Laplace-Delta method as in the
dynamic exDQLM. The AL path sets $\gamma=0$ and uses
$q(\vect{\beta})q(\sigma)\prod_iq(v_i)$ with the same closed-form updates as
Section~\ref{sec:dyn_dqlm}, replacing state smoothing moments by
$\E\{(y_i-\vect{x}_i^\top\vect{\beta})^2\}$.

\paragraph{Algorithm S7 (static AL/exAL LDVB).}
\begin{enumerate}[label=(\arabic*)]
\item Initialize $q(\vect{\beta})$, the latent-variable factors, the scale
factors, and the $(\sigma,\gamma)$ block.
\item Repeat until convergence:
  \begin{enumerate}[label=(\alph*)]
  \item update $q(\vect{\beta})$ using \eqref{eq:static_exal_vb_beta};
  \item update $q(v_i)$ and, for exAL, $q(s_i)$ using
  \eqref{eq:static_exal_vb_vs};
  \item update $q(\sigma)$ for AL or $q(\sigma,\gamma)$ for exAL;
  \item if \code{beta_prior = "rhs_ns"}, update the RHS-NS factors in
  Section~\ref{sec:static_rhs};
  \item monitor moment changes and $\elbo(q)$.
  \end{enumerate}
\item Draw approximate posterior samples when requested.
\end{enumerate}

The static exAL ELBO is computed from the same blocks as the dynamic exDQLM,
with the Gaussian state factor replaced by the Gaussian coefficient factor.
Let
$\bar\ell_\sigma=\E(\log\sigma)$,
$\bar\ell_B=\E(\log B_\gamma)$,
$\bar\ell_\pi=\E\{\log\pigamma(\gamma)\}$,
$m_{v,i}=\E(v_i)$, $m_{1/v,i}=\E(v_i^{-1})$,
$m_{\log v,i}=\E(\log v_i)$, $m_{s,i}=\E(s_i)$, and
$m_{s^2,i}=\E(s_i^2)$. Also define
\begin{equation}
t_i=y_i-\vect{x}_i^\top\vect{m}_\beta,\qquad
q_i=\vect{x}_i^\top\mat{\Sigma}_\beta\vect{x}_i.
\end{equation}
Then
\begin{align}
\elbo(q)
&=
\elbo_{\mathrm{like}}+\elbo_v+\elbo_s+\elbo_\beta
+\elbo_{\sigma\gamma}
+H\{q(\vect{\beta})\}
+\sum_{i=1}^nH\{q(v_i)\}\nonumber\\
&\quad
+\sum_{i=1}^nH\{q(s_i)\}
+H\{q(\sigma,\gamma)\},
\label{eq:static_exal_elbo}
\end{align}
where
\begin{align}
\elbo_{\mathrm{like}}
&=
-\frac{n}{2}\log(2\pi)
-\frac{n}{2}(\bar\ell_B+\bar\ell_\sigma)
-\frac{1}{2}\sum_{i=1}^nm_{\log v,i}\nonumber\\
&\quad
-\frac{1}{2}\sum_{i=1}^n
\left\{
\kappa_1m_{1/v,i}(t_i^2+q_i)
-2\kappa_2t_i+\kappa_3m_{v,i}
\right\}\nonumber\\
&\quad
+\sum_{i=1}^n
\left\{
\kappa_4m_{s,i}m_{1/v,i}t_i
-\kappa_6m_{s,i}
-\frac{1}{2}\kappa_5m_{s^2,i}m_{1/v,i}
\right\},\\
\elbo_v
&=
-n\bar\ell_\sigma-\E(\sigma^{-1})\sum_{i=1}^nm_{v,i},\\
\elbo_s
&=
n\log 2-\frac{n}{2}\log(2\pi)
-\frac{1}{2}\sum_{i=1}^nm_{s^2,i},\\
\elbo_{\sigma\gamma}
&=
a_\sigma\log b_\sigma-\log\Gamma(a_\sigma)
-(a_\sigma+1)\bar\ell_\sigma
-b_\sigma\E(\sigma^{-1})
+\bar\ell_\pi,\\
\elbo_\beta
&=\E_q\{\log p(\vect{\beta})\}.
\label{eq:static_exal_elbo_terms}
\end{align}
For the ridge prior, $\elbo_\beta$ is the Gaussian prior expectation. For the
RHS-NS prior, it is replaced by the RHS-NS contribution in
Section~\ref{sec:static_rhs}. The Gaussian, GIG, truncated-normal, and
Laplace-Delta entropy terms are given by
\eqref{eq:entropy_gaussian}, \eqref{eq:entropy_gig},
\eqref{eq:entropy_tnp}, and \eqref{eq:entropy_ld}, respectively. The AL path
is recovered by setting $\gamma=0$ and dropping the $s_i$ and
$(\sigma,\gamma)$ Laplace-Delta blocks; then $q(\sigma)$ is inverse-gamma and
uses \eqref{eq:entropy_ig}.

\section{Static regression with the RHS-NS prior}
\label{sec:static_rhs}

This section focuses on the \code{rhs_ns} coefficient prior, the
Gibbs-friendly regularized horseshoe variant used for sparse static
regression. It enters the static Gaussian update for $\vect{\beta}$ through a
prior precision matrix. Let
$\mathcal{A}\subseteq\{1,\ldots,p\}$ denote the active shrinkage set; by
default an intercept, if present, is not shrunk. For $j\notin\mathcal{A}$ the
package uses a fixed intercept precision. The direct \code{rhs} path is not
developed here because its log-scale block is nonconjugate; the
\code{rhs_ns} path is the conjugate sparse-prior formulation documented in
this supplement.

\subsection{RHS-NS prior hierarchy}

For $j\in\mathcal{A}$, the \code{rhs_ns} hierarchy is
\begin{align}
\beta_j\mid\tau^2,\lambda_j^2,\zeta^2
&\propto
\N(\beta_j\mid0,\tau^2\lambda_j^2)
\N(\beta_j\mid0,\zeta^2),\\
\lambda_j^2\mid\nu_j &\sim \IG(1/2,1/\nu_j),&
\nu_j&\sim\IG(1/2,1),\\
\tau^2\mid\xi &\sim\IG(1/2,1/\xi),&
\xi&\sim\IG(1/2,1/\tau_0^2),\\
\zeta^2&\sim\IG(a_\zeta,b_\zeta),
\label{eq:rhs_ns_hierarchy}
\end{align}
unless $\zeta^2$ is fixed by the user. The implementation includes the
scale-dependent Gaussian log-normalizing terms shown in the ELBO contribution
below. Thus the proportionality in \eqref{eq:rhs_ns_hierarchy} denotes the
coefficient kernel, while the surrounding scale updates and ELBO use the
corresponding normal-density contributions. The resulting prior precision
contribution for active coefficients is
\begin{equation}
\omega^{\mathrm{prior}}_j=\frac{1}{\tau^2\lambda_j^2}+\frac{1}{\zeta^2}.
\label{eq:rhs_ns_precision}
\end{equation}
Consequently, the $\vect{\beta}$ update in MCMC or VB is obtained from the
static Gaussian update after replacing $\mat{V}_{\beta,0}^{-1}$ by a diagonal precision
matrix with active entries $\omega^{\mathrm{prior}}_j$ or their variational
expectations.

\subsection{RHS-NS MCMC updates}

Conditioning on $\vect{\beta}$, the shrinkage block has closed-form
inverse-gamma updates:
\begin{align}
\lambda_j^2\mid\cdot
&\sim
\IG\left(1,\ \frac{1}{\nu_j}+\frac{\beta_j^2}{2\tau^2}\right),\\
\nu_j\mid\cdot
&\sim
\IG\left(1,\ 1+\frac{1}{\lambda_j^2}\right),\\
\tau^2\mid\cdot
&\sim
\IG\left(\frac{|\mathcal{A}|+1}{2},\
\frac{1}{\xi}+\frac{1}{2}\sum_{j\in\mathcal{A}}\frac{\beta_j^2}{\lambda_j^2}
\right),\\
\xi\mid\cdot
&\sim
\IG\left(1,\ \frac{1}{\tau_0^2}+\frac{1}{\tau^2}\right),\\
\zeta^2\mid\cdot
&\sim
\IG\left(a_\zeta+\frac{|\mathcal{A}|}{2},\
b_\zeta+\frac{1}{2}\sum_{j\in\mathcal{A}}\beta_j^2
\right),
\label{eq:rhs_ns_mcmc}
\end{align}
with the last line omitted when $\zeta^2$ is fixed.
These equations are the conditional targets for the RHS-NS scale block. The
implementation may temporarily freeze or schedule the global-scale update during
configured warmup iterations for numerical stability; this changes only the
update schedule, not the conditional target displayed above.

\subsection{RHS-NS VB updates and ELBO contribution}

The package uses a mean-field approximation over the RHS-NS scale block. For
$j\in\mathcal{A}$,
\begin{align}
q(\lambda_j^2)&=\IG(a_{\lambda j},b_{\lambda j}),&
q(\nu_j)&=\IG(a_{\nu j},b_{\nu j}),\\
q(\tau^2)&=\IG(a_\tau,b_\tau),&
q(\xi)&=\IG(a_\xi,b_\xi),\\
q(\zeta^2)&=\IG(a_\zeta^\star,b_\zeta^\star),
\end{align}
with $a_{\lambda j}=a_{\nu j}=a_\xi=1$ and
$a_\tau=(|\mathcal{A}|+1)/2$. Let
$\bar\beta_j^2=\E_q(\beta_j^2)$. The coordinate updates are
\begin{align}
b_{\lambda j}
&=\frac{1}{2}\bar\beta_j^2\,\E(\tau^{-2})+\E(\nu_j^{-1}),\\
b_{\nu j}
&=1+\E\{(\lambda_j^2)^{-1}\},\\
b_\tau
&=\frac{1}{2}\sum_{j\in\mathcal{A}}\bar\beta_j^2
\E\{(\lambda_j^2)^{-1}\}+\E(\xi^{-1}),\\
b_\xi
&=\frac{1}{\tau_0^2}+\E(\tau^{-2}),\\
a_\zeta^\star
&=a_\zeta+\frac{|\mathcal{A}|}{2},\qquad
b_\zeta^\star=b_\zeta+\frac{1}{2}\sum_{j\in\mathcal{A}}\bar\beta_j^2.
\label{eq:rhs_ns_vb_updates}
\end{align}
The third line is shorthand for
$b_\tau=\frac{1}{2}\sum_{j\in\mathcal{A}}\bar\beta_j^2
\E\{(\lambda_j^2)^{-1}\}+\E(\xi^{-1})$, where the expectation of
$(\lambda_j^2)^{-1}$ is component-specific inside the summation.
As in MCMC, optional warmup scheduling for the $\tau^2,\xi$ block affects when
the coordinate is updated, not the coordinate target itself.

The following term-by-term ELBO contribution matches the package's
\code{rhs_ns} scale block. For $x\sim\IG(a,b)$ under the rate
parameterization used above, define
\begin{align}
m_{\log x}(a,b)&=\log b-\psi(a),&
m_{x^{-1}}(a,b)&=a/b,\\
h_{\IG}(a,b)&=a+\log b+\log\Gamma(a)-(a+1)\psi(a),
\end{align}
where $\psi(\cdot)$ is the digamma function and $h_{\IG}$ is the entropy of an
inverse-gamma variational factor. Let
\begin{align}
\ell_{\lambda j}&=m_{\log x}(a_{\lambda j},b_{\lambda j}),&
r_{\lambda j}&=m_{x^{-1}}(a_{\lambda j},b_{\lambda j}),\\
\ell_{\nu j}&=m_{\log x}(a_{\nu j},b_{\nu j}),&
r_{\nu j}&=m_{x^{-1}}(a_{\nu j},b_{\nu j}),\\
\ell_\tau&=m_{\log x}(a_\tau,b_\tau),&
r_\tau&=m_{x^{-1}}(a_\tau,b_\tau),\\
\ell_\xi&=m_{\log x}(a_\xi,b_\xi),&
r_\xi&=m_{x^{-1}}(a_\xi,b_\xi).
\end{align}
If $\zeta^2$ is random, set
$\ell_\zeta=m_{\log x}(a_\zeta^\star,b_\zeta^\star)$ and
$r_\zeta=m_{x^{-1}}(a_\zeta^\star,b_\zeta^\star)$; if it is fixed at
$\zeta_0^2$, set $\ell_\zeta=\log\zeta_0^2$ and
$r_\zeta=1/\zeta_0^2$.

The coefficient-prior contribution is
\begin{align}
\elbo_{\beta,\mathrm{hs}}
&=
\sum_{j\in\mathcal{A}}
\left\{
-\frac12\log(2\pi)
-\frac12\ell_\tau
-\frac12\ell_{\lambda j}
-\frac12\bar\beta_j^2 r_\tau r_{\lambda j}
\right\},\\
\elbo_{\beta,\mathrm{slab}}
&=
\sum_{j\in\mathcal{A}}
\left\{
-\frac12\log(2\pi)
-\frac12\ell_\zeta
-\frac12\bar\beta_j^2 r_\zeta
\right\}.
\end{align}
The local, global, and slab prior contributions are
\begin{align}
\elbo_{\lambda\mid\nu}
&=
\sum_{j\in\mathcal{A}}
\left\{
-\frac12\ell_{\nu j}
-\log\Gamma(1/2)
-\frac32\ell_{\lambda j}
-r_{\nu j}r_{\lambda j}
\right\},\\
\elbo_{\nu}
&=
\sum_{j\in\mathcal{A}}
\left\{
-\log\Gamma(1/2)
-\frac32\ell_{\nu j}
-r_{\nu j}
\right\},\\
\elbo_{\tau\mid\xi}
&=
-\frac12\ell_\xi
-\log\Gamma(1/2)
-\frac32\ell_\tau
-r_\xi r_\tau,\\
\elbo_{\xi}
&=
\frac12\log(\tau_0^{-2})
-\log\Gamma(1/2)
-\frac32\ell_\xi
-\tau_0^{-2}r_\xi,\\
\elbo_{\zeta}
&=
\begin{cases}
a_\zeta\log b_\zeta-\log\Gamma(a_\zeta)
-(a_\zeta+1)\ell_\zeta-b_\zeta r_\zeta,
& \zeta^2\text{ random},\\
0, & \zeta^2\text{ fixed}.
\end{cases}
\end{align}
The entropy contribution is
\begin{equation}
\elbo_{\mathrm{ent}}
=
\sum_{j\in\mathcal{A}} h_{\IG}(a_{\lambda j},b_{\lambda j})
+\sum_{j\in\mathcal{A}} h_{\IG}(a_{\nu j},b_{\nu j})
+h_{\IG}(a_\tau,b_\tau)
+h_{\IG}(a_\xi,b_\xi)
+\Ind(\zeta^2\text{ random})h_{\IG}(a_\zeta^\star,b_\zeta^\star).
\end{equation}
Finally, when the intercept is not included in $\mathcal{A}$, the package adds
the fixed Gaussian intercept-prior contribution
\begin{equation}
\elbo_{\beta_0}
=
\frac12\{\log(\pi_0)-\log(2\pi)\}
-\frac12\pi_0\bar\beta_0^2,
\end{equation}
where $\pi_0$ is the fixed intercept precision. If the intercept is shrunk,
this term is omitted because the intercept is already represented in the
active RHS-NS block.

Thus the package contribution is
\begin{equation}
\elbo_{\mathrm{rhs\_ns}}
=
\elbo_{\beta,\mathrm{hs}}
+\elbo_{\beta,\mathrm{slab}}
+\elbo_{\lambda\mid\nu}
+\elbo_{\nu}
+\elbo_{\tau\mid\xi}
+\elbo_{\xi}
+\elbo_{\zeta}
+\elbo_{\mathrm{ent}}
+\elbo_{\beta_0}.
\label{eq:rhs_ns_elbo}
\end{equation}
All terms are available in closed form because each scale factor is
inverse-gamma. This is the RHS-NS component of the full static-model ELBO; the
likelihood, latent-variable, and $(\sigma,\gamma)$ terms are added by the
surrounding static AL or exAL variational routine.

\section{Forecasting, diagnostics, and posterior-predictive synthesis}
\label{sec:forecast_diagnostics_synthesis}

\subsection{Forecasting}

For a posterior draw $m$ at forecast origin $\tilde t$, the state forecast
recursion in the main article gives
\begin{equation}
\theta_{\tilde t+k}^{(m)}
\sim
\N\{a_{\tilde t}^{(m)}(k),R_{\tilde t}^{(m)}(k)\},
\qquad k=1,\ldots,K.
\end{equation}
Given a forecasted state draw, posterior predictive draws are generated from
\begin{equation}
Y_{\tilde t+k}^{\mathrm{rep},(m)}
\sim
\exAL_{p_0}\{
\mat{F}_{\tilde t+k}^\top\theta_{\tilde t+k}^{(m)},
\sigma^{(m)},\gamma^{(m)}
\}.
\label{eq:forecast_postpred}
\end{equation}
For the AL/DQLM special case, set $\gamma^{(m)}=0$. The function
\code{exdqlmForecast()} implements this recursion for MCMC and VB fitted
objects; for VB fits, the posterior draws are replaced by draws from the fitted
variational factors.

\subsection{Diagnostics}

The dynamic diagnostics use the MAP one-step-ahead standardized forecast-error
sequence produced by the filtering recursion. If $\widehat f_t$ and
$\widehat q_t$ denote the fitted one-step-ahead location and variance used in
this plug-in sequence, the diagnostic error and its normal-score transform are
\begin{equation}
\widehat e_t = \frac{y_t-\widehat f_t}{\sqrt{\widehat q_t}},
\qquad
\widehat u_t = \Phi(\widehat e_t).
\label{eq:pit_plugin}
\end{equation}
The QQ plot, ACF plot, KL divergence, and standardized forecast-error plot are
computed from $\{\widehat e_t\}$ or $\{\widehat u_t\}$ rather than from
posterior uncertainty bands for residuals.

For predictive draws $y_t^{(1)},\ldots,y_t^{(M)}$, the empirical CRPS estimator
is
\begin{equation}
\widehat{\CRPS}_t
=
\frac{1}{M}\sum_{m=1}^M |y_t^{(m)}-y_t^{\mathrm{obs}}|
-
\frac{1}{2M^2}\sum_{m=1}^M\sum_{\ell=1}^M
|y_t^{(m)}-y_t^{(\ell)}|.
\label{eq:crps_estimator}
\end{equation}
The reported CRPS is the average of \eqref{eq:crps_estimator} over the
diagnostic time points \citep{gneiting2007}. For target quantile $p_0$, the
posterior predictive loss criterion used by the package is
\begin{equation}
\widehat{\pplc}
=
\sum_t
\frac{1}{M}\sum_{m=1}^M
\rho_{p_0}\{y_t^{\mathrm{obs}}-y_t^{\mathrm{rep},(m)}\},
\label{eq:pplc_estimator}
\end{equation}
where $\rho_{p_0}$ is the check-loss function \citep{gelfand1998model}.
The dynamic diagnostic routine returns these summaries. The static diagnostic
routine returns check-loss summaries and reference-quantile errors when a
reference quantile is supplied.

\subsection{Posterior-predictive synthesis}

The function \code{quantileSynthesis()} combines posterior predictive draws
from separately fitted models at ordered levels $p_1<\cdots<p_K$. For each time
point and draw index, it forms the quantile-specific predictive curve, applies
the requested monotonicity correction when fitted quantiles cross, and
interpolates across $p$ to obtain draws from one posterior predictive
distribution. The monotonicity correction uses isotonic adjustment and, when
requested, global monotone rearrangement \citep{barlow1972,chernozhukov2010}.

\paragraph{Algorithm S8 (posterior-predictive synthesis).}
\begin{enumerate}[label=(\arabic*)]
\item Collect posterior predictive draw matrices from fitted models at
$p_1<\cdots<p_K$.
\item For each time point and draw index, form the quantile-specific predictive
curve.
\item Apply the requested monotonicity correction.
\item Interpolate across quantile levels.
\item Return synthesized posterior predictive draws and summaries.
\end{enumerate}

\section{Numerical and ELBO implementation blocks}
\label{sec:numerical_blocks}

\subsection{Reusable moment and entropy formulas}
\label{sec:elbo_building_blocks}

The VB routines monitor
\begin{equation}
\elbo(q)=\E_q\{\log p(\text{complete data})\}+H(q),
\qquad
H(q)=-\E_q\{\log q\}.
\end{equation}
The following moment and entropy formulas are reused by the dynamic and static
algorithms.

If $\vect{x}\sim\N(\vect{m},\mat{V})$ with
$\vect{x}\in\R^d$, then
\begin{equation}
H\{\N(\vect{m},\mat{V})\}
=\frac{1}{2}\{d(1+\log 2\pi)+\log|\mat{V}|\}.
\label{eq:entropy_gaussian}
\end{equation}
If $x\sim\IG(a,b)$ under the parameterization in Section S2, then
\begin{equation}
\E(x^{-1})=\frac{a}{b},\qquad
\E(\log x)=\log b-\psi(a),
\end{equation}
and
\begin{equation}
H\{\IG(a,b)\}
=a+\log b+\log\Gamma(a)-(a+1)\psi(a),
\label{eq:entropy_ig}
\end{equation}
where $\psi(\cdot)$ is the digamma function.

If $x\sim\GIG(\lambda,\chi,\psi_{\mathrm{GIG}})$ and
$z=(\chi\psi_{\mathrm{GIG}})^{1/2}$, let
$m_x=\E(x)$, $m_{1/x}=\E(x^{-1})$, and
$m_{\log x}=\E(\log x)$. Then
\begin{align}
H\{\GIG(\lambda,\chi,\psi_{\mathrm{GIG}})\}
&=
\log\{2K_\lambda(z)\}
-\frac{\lambda}{2}\log\left(\frac{\psi_{\mathrm{GIG}}}{\chi}\right)
-(\lambda-1)m_{\log x} \nonumber\\
&\quad+
\frac{1}{2}\{\chi m_{1/x}+\psi_{\mathrm{GIG}} m_x\}.
\label{eq:entropy_gig}
\end{align}
The moments $m_x$ and $m_{1/x}$ follow from \eqref{eq:gig_moments};
$m_{\log x}$ is the derivative of $\log K_\lambda(z)$ with respect to
$\lambda$ plus $\frac{1}{2}\log(\chi/\psi_{\mathrm{GIG}})$, and is evaluated
numerically in the package.

If $x\sim\TNp(m,V)$, let $a=V^{1/2}$, $z=m/a$, and
$\Lambda(z)=\phi(z)/\Phi(z)$. The entropy is
\begin{equation}
H\{\TNp(m,V)\}
=
\frac{1}{2}\log(2\pi V)+\log\Phi(z)
+\frac{1}{2}\{1-z\Lambda(z)\}.
\label{eq:entropy_tnp}
\end{equation}
This closed form is used for the truncated-normal latent factors; numerically,
the package evaluates the required moments using stable tail-probability
guards.

For exAL VB, the nonconjugate $(\sigma,\gamma)$ block is handled on
transformed coordinates
\begin{equation}
\ell_\sigma=\log\sigma,\qquad
z_\gamma=\logit\left(\frac{\gamma-L}{U-L}\right).
\end{equation}
If $\vect{z}=(z_\gamma,\ell_\sigma)^\top$ has Laplace approximation
$q_z(\vect{z})=\N(\hat{\vect{z}},\mat{\Sigma}_z)$ and
$(\gamma,\sigma)=T(\vect{z})$, then the entropy contribution on the original
scale is approximated as
\begin{equation}
H\{q(\sigma,\gamma)\}
\approx
\frac{1}{2}\{2(1+\log 2\pi)+\log|\mat{\Sigma}_z|\}
+\E_{q_z}\{\log|J_T(\vect{z})|\},
\label{eq:entropy_ld}
\end{equation}
where
\begin{equation}
\log|J_T(\vect{z})|
=\ell_\sigma+\log(U-L)+\log r+\log(1-r),
\qquad r=\expit(z_\gamma).
\end{equation}
The final expectation in \eqref{eq:entropy_ld}, and other smooth expectations
of functions of $(\sigma,\gamma)$, are evaluated by the same second-order
Delta approximation used in the LDVB updates.

\subsection{Bounded univariate slice sampler}
\label{sec:slice}

Let $h(x)$ be an unnormalized log-density on a finite interval $[a,b]$, and let
$x_0\in(a,b)$. The bounded slice update implemented in the package is:
\begin{enumerate}[label=(\arabic*)]
\item Draw $e\sim\Exp(1)$ and set the log-slice height $z=h(x_0)-e$.
\item Draw $u\sim\operatorname{Unif}(0,w)$ and initialize
$L=x_0-u$, $R=x_0+w-u$.
\item Step out left and right by increments of $w$, respecting the bounds
$a$ and $b$, until the interval endpoints lie below the slice or the maximum
number of expansions is reached.
\item Truncate the interval to $[a,b]$.
\item Repeatedly draw $x_1\sim\operatorname{Unif}(L,R)$. If $h(x_1)\ge z$,
accept $x_1$. Otherwise shrink the interval to $[L,x_1]$ if $x_1>x_0$ and to
$[x_1,R]$ otherwise.
\end{enumerate}
For dynamic and static exAL MCMC fits, this sampler is applied to the
one-dimensional $\gamma$ target on $(L,U)$ by default.

\subsection{Laplace-Delta approximation}

For a nonconjugate block $\vect{\eta}\in\R^d$ with log target
$\ell(\vect{\eta})$, let $\hat{\vect{\eta}}$ be a local mode and
$\mat{H}=\nabla^2\ell(\hat{\vect{\eta}})$. If $-\mat{H}$ is positive definite,
the Laplace approximation is
\begin{equation}
q(\vect{\eta})\approx
\N\left(\hat{\vect{\eta}},\ [-\mat{H}]^{-1}\right).
\label{eq:ld_gaussian}
\end{equation}
For a smooth function $r(\vect{\eta})$, the Delta approximation used by LDVB is
\begin{equation}
\E\{r(\vect{\eta})\}
\approx
r(\hat{\vect{\eta}})
+
\frac{1}{2}\tr\left[
\nabla^2 r(\hat{\vect{\eta}})[-\mat{H}]^{-1}
\right].
\label{eq:delta_generic}
\end{equation}
This is used for smooth functions of $(\sigma,\gamma)$ in dynamic and static
exAL models.

\section{Package-to-algorithm map}
\label{sec:package_algorithm_map}

Sections S3--S9 define the targets used by the exported fitting and
post-processing functions. Some routines temporarily hold or schedule selected
blocks during early iterations for numerical stability. These controls affect
update timing only; they do not change the posterior target, full conditional
distributions, or coordinate optima listed in this supplement.

\begin{table}[!htb]
\centering
\begingroup
\footnotesize
\begin{tabularx}{\textwidth}{@{}p{0.30\textwidth}p{0.14\textwidth}>{\raggedright\arraybackslash}X@{}}
\toprule
Function & Section & Algorithmic block \\
\midrule
\texttt{exdqlmMCMC()} & S3--S4 & Dynamic DQLM/exDQLM MCMC; Algorithms S1 and S3. \\
\texttt{exdqlmLDVB()} & S3--S4 & Dynamic DQLM/exDQLM LDVB; Algorithms S2 and S4. \\
\texttt{exdqlmTransferMCMC()} & S5 & Transfer-function MCMC; Algorithm S5 with Algorithm S3. \\
\texttt{exdqlmTransferLDVB()} & S5 & Transfer-function LDVB; Algorithm S5 with Algorithm S4. \\
\texttt{exalStaticMCMC()} & S6--S7 & Static AL/exAL MCMC; Algorithm S6. \\
\texttt{exalStaticLDVB()} & S6--S7 & Static AL/exAL LDVB; Algorithm S7. \\
\texttt{exdqlmForecast()} & S8.1 & Forecast recursion and predictive draws. \\
\texttt{exdqlmDiagnostics()} & S8.2 & Dynamic PIT, CRPS, and pplc diagnostics. \\
\texttt{exalStaticDiagnostics()} & S8.2 & Static check-loss and reference-quantile diagnostics. \\
\texttt{quantileSynthesis()} & S8.3 & Posterior-predictive synthesis; Algorithm S8. \\
\begin{tabular}[t]{@{}l@{}}\texttt{dexal()}, \texttt{pexal()}\\ \texttt{qexal()}, \texttt{rexal()}\end{tabular} & S2 & exAL distribution evaluation and simulation. \\
\texttt{get\_gamma\_bounds()} & S2 & Roots $L,U$ for a fixed $p_0$. \\
\bottomrule
\end{tabularx}
\endgroup
\caption{Mapping between exported \pkg{exdqlm} functions and the algorithmic
blocks derived in this supplement.}
\label{tab:supp_algorithm_map}
\end{table}

\bibliography{bibl}

\end{document}
